---
title: "CFA Project"
author: "Allama Ikbal Sijan"
date: "2024-11-30"
site: bookdown::bookdown_site
output:
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    theme: united
    number_sections: false
    toc_float:
      collapsed: yes
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages upload, message=FALSE, warning=FALSE}
library(tidyverse)
library(psych)
library(lavaan)
library(haven)
library(stringr)
library(semTools)
library(apaTables)
library(bookdown)
```

## Data Cleaning

```{r loading data, warning=FALSE}
# Loading dataset
df <- read_sav("discounting_data.sav")

# Checking for incomplete data
table(df$Finished)

# Keeping only relevant variable
df <- df %>% 
  select(c("ResponseId", starts_with("defer"), starts_with("expedite"), 
           "Age_1", "Race", "Gender"))

# Checking whether the three unfinished datapoints have discounting task completed
na_count <- function(df){
  na_dict <- list()
  
  for (var in colnames(df)){
    total_na <- sum(is.na(df$var))
    na_dict[[var]] = total_na
  }
    
  return(na_dict)
}
print(unlist(na_count(df)))
```
All 500 data points appear to have complete data for the discounting tasks; however, many values are effectively missing, but they are not labeled as NAs.

```{r, echo=FALSE, eval=FALSE}
# Saving the dataset
write.csv(df, "discounting_data.csv", row.names = F)
```

```{r, warning=FALSE}
# Checking datatype of discounting items
class(df$defer1) # the discounting items should be numeric

# identifying the participants that used `$` sign
set_dolla_sign <- list()

for (var in names(df)){
  index_vec <- str_which(df[[var]], "^\\$")
  for (vec in index_vec) {
    set_dolla_sign <- union(set_dolla_sign, vec)
  }
}

paste0("List of participants that used $ before their input ", paste(unlist(set_dolla_sign), collapse = ", "))
```

```{r}
# Removing the $ sign from individual responses
df <- data.frame(lapply(df, function(x) str_remove(x, "^\\$")))

# Uncomment to check if the function worked
# str_which(df$defer1, "^\\$")

# removing the `,` from individual responses as that coerces the value to be NA
df <- data.frame(lapply(df, function(x) str_remove_all(x, ",")))

# Removing `.` at the end of the data as that coerces the value to be NA
df <- data.frame(lapply(df, function(x) str_remove(x, "\\.+$")))
```

```{r}
# Detecting index of text responses (that will be converted to NAs)
text_NAs <- list()

for (var in df %>% select(-ResponseId) %>% names()){
  index_vec <- str_which(df[[var]], "[^0-9.]")
  for (index in index_vec){
    text_NAs <- union(text_NAs, index)
  }
}

paste0("List of participants that had texts in their responses ", paste(unlist(text_NAs), collapse = ", "))
```

```{r, warning=FALSE}
# Converting discount item responses (character to numeric)
new_df <- data.frame(lapply(df[, -which(names(df) == "ResponseId")], as.numeric))
new_df <- new_df %>% 
  add_column(df$ResponseId, .before = 1) %>% 
  rename(ResponseId = `df$ResponseId`)

# Converting 0 and empty cells to NA
new_df <- new_df %>% 
  mutate(across(starts_with("expedite"), ~ if_else(. == 0 | . == "", NA, .)))

df <- new_df
```

## Descriptive Statistics
```{r}
# Continuous variables
describe(df[ , -c(1, 15, 16)])

# Categorical variables
table(df$Gender) # Male = 24%, Female = 73%
table(df$Race) # White = 51%, Black = 17%
```

## Missing Data Analysis
```{r}
# Total missing cells
sum(is.na(df)) # 27 NAs

# Function for complete vs. incomplete data
complete <- function(data) {

  subset_data <- data[, -which(names(data) == "ResponseId")]
  complete_rows <- apply(subset_data, 1, function(row) all(!is.na(row)))
  
  return(as.integer(complete_rows))
}

# Applying the function to the dataset
df$Complete <- complete(df)

table(df$Complete)

## Gender differences between complete vs incomplete cases?
table_gender <- table(df$Gender, df$Complete)
chisq.test(table_gender, simulate.p.value = TRUE, B = 2000)

## Racial differences between complete vs incomplete cases?
table_race <- table(df$Race, df$Complete)
chisq.test(table_race, simulate.p.value = TRUE, B = 2000)

## Age differences between complete vs incomplete cases?
t_test <- t.test(df$Age_1 ~ df$Complete)
t_test # highly significant -> NAR


df %>% 
  group_by(Complete) %>% 
  summarise(M = mean(Age_1, na.rm = T), SD = sd(Age_1, na.rm = T))
```

## Data Transformation
```{r}
# Converting Responses to Discount Rates
df <- df %>% 
  mutate(dr_defer1 = (log(defer1 / 40)) / (1/12),
         dr_defer2 = (log(defer2 / 1500)) / (12/12),
         dr_defer3 = (log(defer3 / 75)) / (6/12),
         dr_defer4 = (log(defer4 / 800)) / (1/12),
         dr_defer5 = (log(defer5 / 1250)) / (6/12),
         dr_defer6 = (log(defer6 / 50)) / (12/12),
         dr_expedite1 = (log(40 / expedite1)) / (1/12),
         dr_expedite2 = (log(1500 / expedite2)) / (12/12),
         dr_expedite3 = (log(75 / expedite3)) / (6/12),
         dr_expedite4 = (log(800 / expedite4)) / (1/12),
         dr_expedite5 = (log(1250 / expedite5)) / (6/12),
         dr_expedite6 = (log(50 / expedite6)) / (12/12),
         )

```

## Outlier Detection

```{r}
# Converting discount rate into z-scores
n <-  1
for (var in colnames(df)[grepl("^dr", colnames(df))]){
  new_var <- paste0("z_score", n)
  df[[new_var]] <- as.vector(scale(df[[var]]))
  n <-  n + 1
}

# Converting participants data with extreme z-values to NAs 

df <- df %>% 
  mutate(dr_defer1 = if_else(abs(z_score1) >= 3, NA, dr_defer1),
         dr_defer2 = if_else(abs(z_score2) >= 3, NA, dr_defer2),
         dr_defer3 = if_else(abs(z_score3) >= 3, NA, dr_defer3),
         dr_defer4 = if_else(abs(z_score4) >= 3, NA, dr_defer4), 
         dr_defer5 = if_else(abs(z_score5) >= 3, NA, dr_defer5), 
         dr_defer6 = if_else(abs(z_score6) >= 3, NA, dr_defer6), 
         dr_expedite1 = if_else(abs(z_score7) >= 3, NA, dr_expedite1), 
         dr_expedite2 = if_else(abs(z_score8) >= 3, NA, dr_expedite2),
         dr_expedite3 = if_else(abs(z_score9) >= 3, NA, dr_expedite3),
         dr_expedite4 = if_else(abs(z_score10) >= 3, NA, dr_expedite4), 
         dr_expedite5 = if_else(abs(z_score11) >= 3, NA, dr_expedite5), 
         dr_expedite6 = if_else(abs(z_score12) >= 3, NA, dr_expedite6)) 

```

```{r}
# Converting all the negative discount rates to NA
df <- df %>% 
  mutate(across(starts_with("dr_"), ~ if_else(. < 0, NA, .)))

# Keeping only the relevant variables
cfa_df <- df %>% 
  select(ResponseId, starts_with("dr_"))
```

```{r, warning=FALSE}
# Function for visualizing outliers

outlier_graph <- function(data, var){
  ggplot(data, aes(x = "", y = data[[var]])) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16, outlier.size = 2) +
  theme_linedraw() +
  labs(title = "Boxplot for discounting rates",
       x = paste0("Item ", var),
       y = "Discounting Rates")
}

outlier_graph(df, "dr_defer2") # change the variable
```

## Assumptions check
```{r}
# Checking for multicollinearity 
lowerCor(df[,2:13], use = "complete.obs", show = TRUE) 
```
The following items have multicollinearity issues - expd3, expd4, exp6

```{r, eval=FALSE}
# Saving the APA correlation table
cor_matrix <- cor(df[ ,2:13], use = "pairwise.complete.obs")

apa.cor.table(
  cor_matrix,
  filename = "cor_table.doc",
  table.number = 1,
  show.conf.interval = FALSE
)
```


## CFA Models
```{r}
# Renaming the variables for easier process
rename_custom <- function(name) {
  name <- sub("^dr_defer(\\d+)$", "def\\1", name)  
  name <- sub("^dr_expedite(\\d+)$", "exp\\1", name)
  return(name)
}

cfa_df <- cfa_df %>% 
  rename_with(rename_custom)

# How many missing cells?
sum(is.na(cfa_df)) # 593 NAs -> Use `FIML` Imputation
```

### One Factor Model
```{r, warning=FALSE}
# Model specification
of_model <- 'discount =~ def1 + def2 + def3 + def4 + def5 + def6 +
                         exp1 + exp2 + exp3 + exp4 + exp5 + exp6'

# Fitting the model
of_model_fit <- cfa(of_model, data = cfa_df, missing = "fiml")

summary(of_model_fit, fit.measures = TRUE, rsq = TRUE, standardized = TRUE)

reliability(of_model_fit)
```

#### Model Visualizations
```{r}
# Plot for the one factor model
semPlot::semPaths(of_model_fit, 
                  layout = "tree", style = "lisrel", what = "col", whatLabels = "stand")
```
#### Checking Linearity Assumptions
```{r}
# checking linearity assumptions for one factor model
tf_fs <- data.frame(lavPredict(tf_model_fit, method = "regression"))

# Reshaping the data
plot_data <- cfa_df %>%
  select(def1:def6, exp1:exp6) %>%  # Select the relevant columns
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>% 
  mutate(Discount_Rate = rep(tf_fs$discount, each = 12))

# Create the plot using ggplot2
ggplot(plot_data, aes(x = Discount_Rate, y = Value)) +
  geom_point(size = 0.5, color = "blue", alpha = 0.5) + 
  facet_wrap(~Variable, scales = "free_y") +  
  labs(x = "Discount", y = "Value", title = "Facet Wrap of Discount vs Variables") +
  theme_minimal()
```

### Two Factor Model

```{r, warning=FALSE}
# Model Specification
tf_model <- 'defer =~ def1 + def2 + def3 + def4 + def5 + def6
          expedite =~ exp1 + exp2 + exp3 + exp4 + exp5 + exp6'

# Fitting the Model
tf_model_fit <- cfa(tf_model, data = cfa_df, missing = "fiml")
summary(tf_model_fit, rsq = TRUE, fit.measures = TRUE, standardized = TRUE)

# Checking validity and reliability
semTools::reliability(tf_model_fit)
```

#### Model Visualization
```{r}
# Plot for the two factor model
semPlot::semPaths(tf_model_fit, 
                  layout = "tree", style = "lisrel", what = "col", whatLabels = "stand")
```
#### Checking Linearity Assumptions
```{r, warning=FALSE}
# Checking linearity assumptions for two-factor model
tf_fs <- data.frame(lavPredict(tf_model_fit, method = "regression"))

# Reshaping the data
plot_data <- cfa_df %>%
  select(def1:def6, exp1:exp6) %>%  # Select the relevant observed variables
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  mutate(
    defer = rep(tf_fs$defer, each = 12),
    expedite = rep(tf_fs$expedite, each = 12)
  )

# Create the plot for Factor 1
plot_data_def <- plot_data %>% 
  filter(Variable %in% c("def1", "def2", "def3", "def4", "def5", "def6"))

ggplot(plot_data_def, aes(x = defer, y = Value)) +
  geom_point(size = 0.5, color = "blue", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free_y") +
  labs(x = "Defer Score", y = "Value", title = "Facet Wrap of Factor 1 vs Variables") +
  theme_minimal()

# Create the plot for Factor 2
plot_data_exp <- plot_data %>% 
  filter(Variable %in% c("exp1", "exp2", "exp3", "exp4", "exp5", "exp6"))

ggplot(plot_data_exp, aes(x = expedite, y = Value)) +
  geom_point(size = 0.5, color = "red", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free_y") +
  labs(x = "Expedite Score", y = "Value", title = "Facet Wrap of Factor 2 vs Variables") +
  theme_minimal()
```

### Four Factor Model

```{r, warning=FALSE}
# Model specification
ff_model <- 'defer_small =~ def1 + def3 + def6
             defer_large =~ def2 + def4 + def5
             expedite_small =~ exp1 + exp3 + exp6
             expedite_large =~ exp2 + exp4 + exp5'

# Fitting the model
ff_model_fit <- cfa(ff_model, data = cfa_df, missing = "fiml")
summary(ff_model_fit, fit.measures = TRUE, rsq = TRUE, standardized = TRUE)

# Checking validity and reliability
semTools::reliability(ff_model_fit)
```

#### Model Visualization
```{r}
# Plot for the four factor model
semPlot::semPaths(ff_model_fit, 
                  layout = "tree", style = "lisrel", what = "col", whatLabels = "stand")
```
#### Checking Linearity Assumptions
```{r, warning=FALSE}
# Checking linearity assumptions for four-factor model
ff_fs <- data.frame(lavPredict(ff_model_fit, method = "regression"))

# Reshaping the data
plot_data <- cfa_df %>%
  select(def1:def6, exp1:exp6) %>%  # Select the relevant observed variables
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  mutate(
    defer_small = rep(ff_fs$defer_small, each = 12),
    defer_large = rep(ff_fs$defer_large, each = 12),
    exp_small = rep(ff_fs$expedite_small, each = 12),
    exp_large = rep(ff_fs$expedite_large, each = 12),
  )

# Create the plot for Factor 1
plot_data_def_s <- plot_data %>% 
  filter(Variable %in% c("def1", "def3", "def6"))

ggplot(plot_data_def_s, aes(x = defer_small, y = Value)) +
  geom_point(size = 0.5, color = "lightblue", alpha = 0.5, nrow = 2) +
  facet_wrap(~Variable, scales = "free_y") +
  labs(x = "Defer Score", y = "Value", title = "Facet Wrap of Factor 1 vs Variables") +
  theme_minimal()

# Create the plot for Factor 2
plot_data_def_l <- plot_data %>% 
  filter(Variable %in% c("def2", "def4", "def5"))

ggplot(plot_data_def_l, aes(x = defer_large, y = Value)) +
  geom_point(size = 0.5, color = "blue", alpha = 0.5, nrow = 2) +
  facet_wrap(~Variable, scales = "free_y") +
  labs(x = "Defer Score", y = "Value", title = "Facet Wrap of Factor 1 vs Variables") +
  theme_minimal()

# Create the plot for Factor 3
plot_data_exp_s <- plot_data %>% 
  filter(Variable %in% c("exp1", "exp3", "exp6"))

ggplot(plot_data_exp_s, aes(x = exp_small, y = Value)) +
  geom_point(size = 0.5, color = "darkorange", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free_y", nrow = 2) +
  labs(x = "Expedite Score", y = "Value", title = "Facet Wrap of Factor 2 vs Variables") +
  theme_minimal()

# Create the plot for Factor 4
plot_data_exp_l <- plot_data %>% 
  filter(Variable %in% c("exp2", "exp4", "exp5"))

ggplot(plot_data_exp_l, aes(x = exp_large, y = Value)) +
  geom_point(size = 0.5, color = "red", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free_y", nrow = 2) +
  labs(x = "Expedite Score", y = "Value", title = "Facet Wrap of Factor 2 vs Variables") +
  theme_minimal()
```

## Likelihood Ratio Test
```{r}
# Comparing model one-factor & two-factor model
lavTestLRT(of_model_fit, tf_model_fit)

# Comparing model one-factor & four-factor model
lavTestLRT(of_model_fit, ff_model_fit)

# Comparing model two-factor & four-factor model
lavTestLRT(tf_model_fit, ff_model_fit)
```


