---
title: "CFA_project"
author: "Allama Ikbal Sijan"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages upload, message=FALSE, warning=FALSE}
library(tidyverse)
library(psych)
library(lavaan)
library(haven)
library(stringr)
```

## Data Cleaning

```{r loading data, warning=FALSE}
df <- read_sav("discounting_data.sav")

# Any incomplete data
table(df$Finished)

# only relevant variable
df <- df %>% 
  select(c("ResponseId", starts_with("defer"), starts_with("expedite"), 
           "Age_1", "Race", "Gender"))

# 3 incomplete rows but did they complete the discounting task?
na_count <- function(df){
  na_dict <- list()
  
  for (var in colnames(df)){
    total_na <- sum(is.na(df$var))
    na_dict[[var]] = total_na
  }
    
  return(na_dict)
}

print(unlist(na_count(df)))
```
All 500 data points appear to have complete data for the discounting tasks; however, many values are effectively missing, but they are not labeled as NAs.

```{r}
class(df$defer1) # the discounting items are character types when they should be numeric

# identifying the participants that used `$` sign
set_dolla_sign <- list()

for (var in names(df)){
  index_vec <- str_which(df[[var]], "^\\$")
  for (vec in index_vec) {
    set_dolla_sign <- union(set_dolla_sign, vec)
  }
}

paste0("List of participants that used $ before their input ", paste(unlist(set_dolla_sign), collapse = ", "))
```
```{r}
# removing the $ sign from individual responses
df <- data.frame(lapply(df, function(x) str_remove(x, "^\\$")))

# checking if the function worked
str_which(df$defer1, "^\\$")

# removing the (,) from individual responses as that coerces the value to be NA
df <- data.frame(lapply(df, function(x) str_remove_all(x, ",")))

# also noticed some instances of (.) at the end of the data which coerces into NAs, thus removing those as well
df <- data.frame(lapply(df, function(x) str_remove(x, "\\.+$")))
```

```{r, warning=FALSE}
# detecting index of text responses (that will be converted to NAs)
text_NAs <- list()

for (var in df %>% select(-ResponseId) %>% names()){
  index_vec <- str_which(df[[var]], "[^0-9.]")
  for (index in index_vec){
    text_NAs <- union(text_NAs, index)
  }
}

paste0("List of participants that had texts in their responses ", paste(unlist(text_NAs), collapse = ", "))
```
```{r}
# Character to Numeric
new_df <- data.frame(lapply(df[, -which(names(df) == "ResponseId")], as.numeric))
new_df <- new_df %>% 
  add_column(df$ResponseId, .before = 1) %>% 
  rename(ResponseId = `df$ResponseId`)

# Checks off, it worked
new_df[unlist(text_NAs),]

# converting 0 and empty cells to NA
new_df <- new_df %>% 
  mutate(across(starts_with("expedite"), ~ if_else(. == 0 | . == "", NA, .)))

df <- new_df
```
## Are the missing data random or not?
```{r}
sum(is.na(df)) # 27 NAs

complete <- function(data) {
  # Exclude the "ResponseId" column
  subset_data <- data[, -which(names(data) == "ResponseId")]
  
  # Check each row for NAs and return 1 if complete, 0 otherwise
  complete_rows <- apply(subset_data, 1, function(row) all(!is.na(row)))
  
  return(as.integer(complete_rows))
}

df$Complete <- complete(df)

## differs based on Gender
table_gender <- table(df$Gender, df$Complete)
chisq.test(table_gender) # highly significant -> NAR

## differs based on Race
table_race <- table(df$Race, df$Complete)
chisq.test(table_race) # highly significant -> NAR

## differs based on age
t_test <- t.test(df$Age_1 ~ df$Complete)
t_test # highly significant -> NAR

"""
new_df %>% 
  group_by(Complete) %>% 
  summarise(M = mean(AGE, na.rm = T), SD = sd(AGE, na.rm = T))

cohensD(Age_1 ~ Complete, data = df)
"""
```

## Converting Responses to Discount Rates

```{r}
df <- df %>% 
  mutate(dr_defer1 = (log(defer1 / 40)) / (1/12),
         dr_defer2 = (log(defer2 / 1500)) / (12/12),
         dr_defer3 = (log(defer3 / 75)) / (6/12),
         dr_defer4 = (log(defer4 / 800)) / (1/12),
         dr_defer5 = (log(defer5 / 1250)) / (6/12),
         dr_defer6 = (log(defer6 / 50)) / (12/12),
         dr_expedite1 = (log(40 / expedite1)) / (1/12),
         dr_expedite2 = (log(1500 / expedite2)) / (12/12),
         dr_expedite3 = (log(75 / expedite3)) / (6/12),
         dr_expedite4 = (log(800 / expedite4)) / (1/12),
         dr_expedite5 = (log(1250 / expedite5)) / (6/12),
         dr_expedite6 = (log(50 / expedite6)) / (12/12),
         )

```

### Remove this code section
```{r, eval=FALSE}
"""str_which(df$defer3, "^\\$")
test_df <- data.frame(a = 1:3, b = 4:6)
test_df

# Apply to each value in each column
new_df <- data.frame(apply(test_df, 2, function(x) x * 2)) 
data.frame(lapply(test_df, function(x) x* 2))

# row-wise operations
apply(test_df, 1, function(x) sum(x)) 
data.frame(sapply(test_df, function(x) x* 2))
"""
```

##Outlier Detection

```{r}
# converting discount rate into z-scores
n <-  1
for (var in colnames(df)[grepl("^dr", colnames(df))]){
  new_var <- paste0("z_score", n)
  df[[new_var]] <- as.vector(scale(df[[var]]))
  n <-  n + 1
}

# converting participants data with extreme z-values to NAs 

df <- df %>% 
  mutate(dr_defer1 = if_else(abs(z_score1) >= 3, NA, dr_defer1),
         dr_defer2 = if_else(abs(z_score2) >= 3, NA, dr_defer2),
         dr_defer3 = if_else(abs(z_score3) >= 3, NA, dr_defer3),
         dr_defer4 = if_else(abs(z_score4) >= 3, NA, dr_defer4), 
         dr_defer5 = if_else(abs(z_score5) >= 3, NA, dr_defer5), 
         dr_defer6 = if_else(abs(z_score6) >= 3, NA, dr_defer6), 
         dr_expedite1 = if_else(abs(z_score7) >= 3, NA, dr_expedite1), 
         dr_expedite2 = if_else(abs(z_score8) >= 3, NA, dr_expedite2),
         dr_expedite3 = if_else(abs(z_score9) >= 3, NA, dr_expedite3),
         dr_expedite4 = if_else(abs(z_score10) >= 3, NA, dr_expedite4), 
         dr_expedite5 = if_else(abs(z_score11) >= 3, NA, dr_expedite5), 
         dr_expedite6 = if_else(abs(z_score12) >= 3, NA, dr_expedite6)) 

```

```{r}
# converting all the negative discount rates to NA
df <- df %>% 
  mutate(across(starts_with("dr_"), ~ if_else(. < 0, NA, .)))

# only the relevant variables
cfa_df <- df %>% 
  select(ResponseId, starts_with("dr_"))
```

```{r}
# function to check for outliers using boxplot

outlier_graph <- function(data, var){
  ggplot(data, aes(x = "", y = data[[var]])) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16, outlier.size = 2) +
  theme_linedraw() +
  labs(title = "Boxplot for discounting rates",
       x = paste0("Item ", var),
       y = "Discounting Rates")
}

outlier_graph(cfa_df, "dr_defer1")
```
## CFA Models

Once the discount rate columns have been created and cleaned they can be supplied to CFA models in lavaan. Three that are of interest are:

(1) A one-factor model with all 12 discount rate columns loading onto a single factor.

### One Factor Model
```{r}
# renaming the variables for easier process
rename_custom <- function(name) {
  name <- sub("^dr_defer(\\d+)$", "def\\1", name)  
  name <- sub("^dr_expedite(\\d+)$", "exp\\1", name)
  return(name)
}

cfa_df <- cfa_df %>% 
  rename_with(rename_custom)
```

```{r}
# missing data?
sum(is.na(cfa_df)) # 593 NAs

# model specification
of_model <- 'discount =~ def1 + def2 + def3 + def4 + def5 + def6 +
                         exp1 + exp2 + exp3 + exp4 + exp5 + exp6'

# fitting the model
of_model_fit <- cfa(of_model, data = cfa_df, missing = "fiml")

summary(of_model_fit, fit.measures = TRUE, rsq = TRUE)

"Chi-square <- significant [not a good fit]
CFI = 0.724 <- not a good fit
RMSEA = 0.239 <- not a good fit"
```
### Two Factor Model

```{r}
tf_model <- 'defer =~ def1 + def2 + def3 + def4 + def5 + def6
          expedite =~ exp1 + exp2 + exp3 + exp4 + exp5 + exp6'

tf_model_fit <- cfa(tf_model, data = cfa_df)
summary(tf_model_fit, rsq = TRUE, fit.measures = TRUE)

"Chi-square <- significant [not a good fit]
CFI = 0.935 <- Almost a good fit
RMSEA = 0.127 <- not a good fit"
```

### Four Factor Model
(3) For both defer and expedited items, three of the provided rewards were small in magnitude (< $100) and three were larger in magnitude (>$800). A four-factor model could look at defer_small, defer_large, expedite_small, and expedite_large.

```{r}
ff_model <- 'defer_small =~ def1 + def3 + def6
             defer_large =~ def2 + def4 + def5
             expedite_small =~ exp1 + exp3 + exp6
             expedite_large =~ exp2 + exp4 + exp5'

ff_model_fit <- cfa(ff_model, data = cfa_df)

summary(ff_model_fit, fit.measures = TRUE, rsq = TRUE, standardized = TRUE)

"Chi-square <- significant [not a good fit]
CFI = 0.957 <- Good fit
RMSEA = 0.109 <- not a good fit"
```

Lavaan likelihood ratio tests can then be performed to compare model fits of these nested models (as well as inspection of other global and local fit indices).
```{r}
# Comparing model one-factor & two-factor model
lavTestLRT(of_model_fit, tf_model_fit)

# Comparing model one-factor & four-factor model
lavTestLRT(of_model_fit, ff_model_fit)

# Comparing model two-factor & four-factor model
lavTestLRT(tf_model_fit, ff_model_fit)
```


